Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-03-09 paper.tex
reorganize

  In this paper, we show their correspondence and how to syntactically transform small-step 
  abstracting abstract machines into big-step abstracting definitional 
  interpreters.
  The transformations include fusing, disentangling, refunctionalizing 
  and monadification (or un-CPS to direct-style with delimited continuation). 
  All of them properly handle the 
  collecting semantics and the non-determinism of abstract interpretation. 
  After each transformation, we also obtain an intermediary form of 
  abstract interpreter.
  Remarkably, refunctionalization sequentializes the abstract evaluation.
  And follow the idea that evaluation contexts are defunctionalized 
  continuations, we reveal how precise call/return match in control-flow 
  analysis can be obtained by refunctionalizing a small-step abstract interpreter.
  as well as explain how it been lost by defunctionalizing the higher-order
  continuations into an firts-order data type.

@@

\item Further, we show that defunctionalization and refunctionalization
  of abstract interpreters play important roles for call stack of analyzed language.
  Given this insight, we reveal that why small-step AAM has no pushdown
  control-flow property (if without extra effort), as well as explain
  that why ADI naturally extends pushdown control-flow property from its
  defining language.

@@

Traditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use administrative normal form (ANF) which is a direct-style intermediate representation 
as target for clarity but without losing simplicity and generality.
Although we just show the core calculus language, but it can be easily extended
to support recursive bindings (such as $letrec$), conditionals, primitive types and 
operations on primitive types. These cases would be trival or simple, so we omit them
in the paper.
The transformations we will show in the rest of this paper
also work on abstract machines for plain, direct-style $\lambda$-calculus languages.
TODO: add letrec?

@@

\begin{lstlisting}
e $\in$ Exp ::= ae
          | (let ([x (f ae)]) e)
ae $\in$ AExp ::= x            
            | lam
lam $\in$ Lam ::= (lambda (x) e)
x ... are variable names
\end{lstlisting}

In ANF, all the function applications must be administrated in a @let@ expression,
and then bounded to a variable name under current environment.
We can use @letrec@ to create mutually recursive function bindings. But @letrec@
only accepts atomic expressions that appear at the right-hand side of a binding.
The atomic expressions $ae$ are either a variable, or a literal @lambda@ term, which
both can be evaluated in a single step.

The abstract syntax can be straightforwardly described in Scala as follows:

@@

\begin{lstlisting}
def atomicEval(e: Expr, env: Env, store: Store): Storable = e match {
  case Var(x) => store(env(x))
  case lam@Lam(x, body) => Clos(lam, env)
}
def alloc(store: Store): Addr = store.keys.size + 1
def isAtomic(e: Expr): Boolean = e.isInstanceOf[Var] || e.isInstanceOf[Lam]
\end{lstlisting}

@@

To run (analyze) a program, as CESK machine, we use function @inject@ to construct
the initial state given a program. 
Note that the initial continuation store has a built-in mapping that maps continuation address
@Halt@ to empty set of continuations.
We also provide an empty program contour as initial time.
\begin{lstlisting}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())
\end{lstlisting}

However, comparing with concrete CESK machine, the
@drive@ function performs collecting semantics instead of valuation
semantics. That is, for the purpose of analyzing programs, the function @drive@
collects all the intermediate machine states as the program is abstractly executing.
The following code shows a variant of work list algorithm that always applies @step@
function to the first element @hd@ of the work list @todo@ if @hd@
is unseen, then adds the result of @step@ to the rest of work list and in 
the meanwhile adds @hd@ to the explored states set.
If the work list is empty, the function just returns the set of reachable states up to now.

@@

Finally, a user could use function @analyze@ to obtain all reachable states 
for a given program.

@@

@@

We start from the AAM with unbounded stack.
In fact, since later we will eliminate the call stack and transform to
higher-order continuations, it does not matter what kind of AAM we start from.

@@

In recent years, there are significant efforts \cite{vardoulakis2010cfa2, earl2012introspective, 
gilray2016pushdown, johnson2015abstracting} to achieve precise call/return 
match based on small-step AAM which is what we described in section \ref{aam}.

CFA2 is the first solution that solves the return-flows problem\cite{vardoulakis2010cfa2}.
But CFA2 has several drawbacks: it works only on continuation-passing style programs, 
and only allows monovariant analysis, in addition with an exponential time complexity.
TODO: summarize PDCFA, AAC, P4F

@@

\textbf{Pushdown for Free.}
\citeauthor{gilray2016pushdown} proposed a continuation address allocator for 
small-step abstracting abstract machines that is both simple to implement and
computational inexpensive.
The only change in code is when allocating a continuation address, not only
keep track of the entry-point expression of callee, but also hold the target environment.
The target environment is the binding environment extended from closure's environment
(namely @val new_env = env_c + (v -> baddr)@). All the rest things remain the same.

@@

Notably, this change only causes constant-factor increase in time complexity of analysis
if the store is widened.

\textbf{Abstracting Definitional Interpreters.}
On the other side, with the insight from Reynold's seminal paper 
\emph{Definitional interpreters for higher-order programming languages} \cite{reynolds1972definitional}
that the defined language will inherit properties
from the defining language, \citeauthor{darais2017abstracting} construct abstracting 
definitional interpreters that automatically inherit the pushdown control-flow 
property from its defining language.

TODO: connection to this paper.
TODO: although they establish pushdown control-flow analysis in different ways, 
is there a correspondence between different methods?

>>>>

  In this paper, we show their correspondence and how to syntactically transform small-step 
  abstracting abstract machines into big-step abstracting definitional 
  interpreters.
  The transformations include fusing, disentangling, refunctionalizing 
  and monadification (or un-CPS to direct-style with delimited continuations). 
  All of them properly handle the 
  collecting semantics and the non-determinism of abstract interpretation. 
  After each transformation, we also obtain an intermediary form of 
  abstract interpreter.
  Remarkably, refunctionalization sequentializes the abstract evaluation.
  And follow the idea that evaluation contexts are defunctionalized 
  continuations, we reveal how precise call/return match in control-flow 
  analysis can be obtained by refunctionalizing a small-step abstract interpreter.
  as well as explain how it been lost by defunctionalizing the higher-order
  continuations into an firts-order data type.

@@

\item Further, we show that defunctionalization and refunctionalization
  of abstract interpreters play important roles for call stack of analyzed language.
  Given this insight, we reveal that why small-step AAM has no pushdown
  control-flow property (if without extra efforts), as well as explain
  that why ADI naturally extends pushdown control-flow property from its
  defining language.

@@

Traditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use administrative normal form (ANF) which is a direct-style intermediate representation 
as target for clarity but without losing simplicity and generality.
Although we just show the core calculus language, but it can be easily extended
to support recursive bindings (such as @letrec@), conditionals, primitive types and 
operations on primitive types. These cases would be trival or simple to implement, 
so we omit them in the paper.
The transformations we will show in the rest of this paper
also work on abstract machines for plain direct-style $\lambda$-calculus languages.

@@

\begin{lstlisting}
e $\in$ Exp ::= ae
          | (let ([x (ae ae)]) e)
ae $\in$ AExp ::= x            
            | lam
lam $\in$ Lam ::= (lambda (x) e)
x ... are variable names
\end{lstlisting}

In ANF, an expression is an atomic expression or a @let@ expression.
The restriction is all the function applications must be administrated in a @let@ expression,
and then bounded to a variable name under the current environment.
Both the operator and operand of function applications should be an atomic expression.
The atomic expressions $ae$ are either a variable, or a literal @lambda@ term, which
both can be evaluated in a single step.

The abstract syntax can be straightforwardly described in Scala as follows:
TODO: update the abstract syntax.

@@

\begin{lstlisting}
def atomicEval(e: Expr, env: Env, store: Store): Storable = e match {
  case Var(x) => store(env(x))
  case lam @ Lam(x, body) => Clos(lam, env)
}
def alloc(store: Store): Addr = store.keys.size + 1
def isAtomic(e: Expr): Boolean = e.isInstanceOf[Var] || e.isInstanceOf[Lam]
\end{lstlisting}

@@

To run (analyze) a program, as CESK machines, we first use function @inject@ to construct
the initial state given the program. 
Note that the initial continuation store has a built-in mapping that maps continuation address
@Halt@ to empty set of continuations.
We also provide an empty program contour as initial time.
\begin{lstlisting}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())
\end{lstlisting}

However, comparing with concrete CESK machine, the
@drive@ function performs collecting semantics instead of valuation
semantics. That is, for the purpose of analyzing programs, the function @drive@
collects all the intermediate machine states as the program is abstractly executing.
The following code shows a variant of work list algorithm to find the fixed-point 
of set of states.
Function @drive@ always applies function @step@ to the head element @hd@ of 
the work list @todo@ if @hd@ is unseen, then inserts the result of @step@ to 
the rest of work list; and in the meanwhile adds @hd@ to the explored states set.
If the work list is empty, the @drive@ just returns the set of reachable states 
up to now.

@@

Finally, a user could use function @analyze@ to obtain all reachable states 
as a finite set for a given program.

@@

\subsection{Pushdown Machine}

AAM with unbounded stack

@@

In last section, we showed that by keeping an unabstracted stack in the state 
space we can recover the precise call/return match. In this section, we start
our first transformation: \textit{fusing} an AAM with unbounded stack.

In fact, later we will eliminate the stack and transform it to continuations
represented by higher-order functions.
Because of transforming to higher-order functions forces us to sequentialize the execution order,
it does not matter what kind of AAM we start from. 
We choose to start from AAM with unbounded stack just because TODO.

@@

@@

\textbf{Pushdown Control-flow Analysis}
In recent years, there are significant efforts \cite{vardoulakis2010cfa2, earl2012introspective, 
gilray2016pushdown, johnson2015abstracting} to achieve precise call/return 
match based on small-step AAM which is what we described in section \ref{aam}.

CFA2 is the first solution that solves the return-flows problem\cite{vardoulakis2010cfa2}.
But CFA2 has several drawbacks: it works on only continuation-passing style programs, 
and not supports polyvariant analysis, in addition with an exponential time complexity.
TODO: summarize PDCFA, AAC

\citeauthor{gilray2016pushdown} proposed a continuation address allocator for 
small-step abstracting abstract machines that is both simple to implement and
computational inexpensive. 
Based on the AAM we presented in Section \ref{aam}, the only change in code is when 
allocating a continuation address, not only
keep track of the entry-point expression of callee, but also hold the target environment.
The target environment is the binding environment extended from closure's environment
(namely @val new_env = env_c + (v -> baddr)@). All the rest things remain the same.
The following code shows the implementation:

@@

Notably, this change only causes constant-factor increase in time complexity of 
the analysis if the store is widened.

TOOD: why only needs the target env. correspondence?

\textbf{Abstracting Definitional Interpreters.}
On the other side of small-step AAM, with the insight from Reynold's seminal paper 
\emph{Definitional interpreters for higher-order programming languages} \cite{reynolds1972definitional}
that the defined language will inherit properties from the defining language, 
\citeauthor{darais2017abstracting} construct abstracting 
definitional interpreters that automatically inherit the pushdown control-flow 
property from its defining language.

\citeauthor{darais2017abstracting}'s abstracting definitional interpreters is written
in monadic style.

TODO: connection to this paper.
TODO: delimited control vs monad. correspondence since the very beginning time.
TODO: although they establish pushdown control-flow analysis in different ways, 
is there a correspondence between different methods?

\textbf{Defunctionalization and Refunctionalization.}
